---
title: "Horror Movie Visualizations"
subtitle: "Proposal"
name: "Viz-Villains: Chris Liang, Kevin Ordet, Chris Tsai, Jason Zhang"
format: html
editor: visual
---


```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(patchwork)
library(dplyr)
library(knitr)
```

```{r read-data, message = F}
horror_movies <- read.csv("data/horror_movies.csv")
```


# Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

The data include `r nrow(horror_movies)` entries, with `r ncol(horror_movies)` variables (e.g. the movie name, release date, runtime, etc.). Each entry represents a horror movie released starting from the 1950s. The data were extracted from [The Movie Database](https://www.themoviedb.org) via tmdb API using R `httr` .

We chose the data because it had a healthy amount of both numerical and categorical variables that we thought could produce a variety of visualizations. Additionally, none of us are horror movie fans, so we thought this exploration into a genre we are unfamiliar with would be an enlightening experience.

# Questions

How is the popularity of a released horror movie impacted by the environment that it was released in?

Are there certain sub-genres of horror that perform better than others?

# Analysis plan

We know the first question is vague and it is purposely so, but we hope to make it more clear what we're trying to answer via our analysis and data visualizations. Our first plan of action is to create a visualization that will look at how the popularity of horror movies have changed overtime. We will likely use the popularity and release_date variables for this visualization. Another visualization will look at the same thing, but group by whether the movie is low budget or high budget (have high budget horror movies increased in popularity over time?). We will have to mutate a new categorical variable and add it to the dataset that determines whether the movie is high or low budget. Along with that newly mutate variable we will use the popularity and release_date variable. Third, we will want to look at how the popularity of horror movies increases or decreases after a high profile horror movie is released (either a highly popular, or high budget film). We will use the variables popularity, release_date, and budget for this analysis.


A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
