---
title: "Horror Movie Visualizations"
subtitle: "Proposal"
name: "Viz-Villains: Chris Liang, Kevin Ordet, Chris Tsai, Jason Zhang"
format: html
editor: visual
---


```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(patchwork)
library(dplyr)
library(knitr)
```

```{r read-data, message = F}
horror_movies <- read.csv("data/horror_movies.csv")
```


# Dataset

The data include `r nrow(horror_movies)` entries, with `r ncol(horror_movies)` variables (e.g. the movie name, release date, runtime, etc.). Each entry represents a horror movie released starting from the 1950s. The data were extracted from [The Movie Database](https://www.themoviedb.org) via tmdb API using R `httr` .

We chose the data because it had a healthy amount of both numerical and categorical variables that we thought could produce a variety of visualizations. Additionally, none of us are horror movie fans, so we thought this exploration into a genre we are unfamiliar with would be an enlightening experience.

# Questions

How is the popularity of a released horror movie impacted by the environment that it was released in?

Are there certain sub-genres of horror that perform better than others?

# Analysis plan

We know the first question is vague and it is purposely so, but we hope to make it more clear what we're trying to answer via our analysis and data visualizations. When we say "environment", we mean the context of time of release, as the year of release is linked to different economic environments, different budgets for entertainment, different social norms and audience reception to horror, etc. Our first plan of action is to create a visualization that will look at how the popularity of horror movies have changed overtime. We will likely use the popularity and release_date variables for this visualization. Another visualization will look at the same thing, but group by whether the movie is low budget or high budget (have high budget horror movies increased in popularity over time?). We will have to mutate a new categorical variable and add it to the dataset that determines whether the movie is high or low budget. Along with that newly mutate variable we will use the popularity and release_date variable. Third, we will want to look at how the popularity of horror movies increases or decreases after a high profile horror movie is released (either a highly popular, or high budget film). We will use the variables popularity, release_date, and budget for this analysis. Our last rough idea of a visualization we might want to create is one that looks at how the popularity of foreign language horror films has increased or decreased over time.


For the second question, we aim to look at the variable genre_names, which includes the genres that the movie falls into. Horror, for obvious reasons, is included in the value for the observations of this variable. On top of that, many observations have extra tagged genres, such as drama, thriller, etc. That variable is currently of type character (a string in R), so we can use a delimiter to separate it into different variables for each sub genre. We intend to make indicator variables: beyond horror, it will be Thriller: 0 or 1, 0 for thriller is not included in the extra genre taglines, 1 for if it is. We'll take the top several genres to make indicator variables and discard the rest that have very few tagged -- let's say "Adventure" is too rare a subgenre taged in genre_names, we'll just make indicator variables for the top X occurring subgenres. Afterwards, we can treat these indicator variables as binary categorical variables and create visualizations to compare the indicator variables against reception metrics such as popularity. Using these binary indicator variables can also help us create linear models with binary predictors -- does having a certain subgenre correlate with better reception? We can also mutate genre_names to be a categorical variable with multiple levels, and create visualizations with bar charts, stacked bar charts, etc. all using reception metrics as the basis of comparison.

